Reproduction frequency and offspring survival decline in elephant seals past prime age

### 2024-07-19 paper/paperv3.qmd
missing obs (age 4 on) and quad/break aic

TODO: fix sample size for bioage=0

@@

TODO: include from age 4 on

@@

## Are we missing animals more or less with age?
missing_seals <- sealdat %>% 
  # Need to see a seal 3+ years for calculation to make sense
  group_by(animalID) %>% 
  mutate(n_years = n_distinct(year)) %>% 
  ungroup() %>% 
  filter(n_years >= 3,
         age <= 20) %>% 
  # Fill in non-resighted years
  group_by(animalID) %>% 
  arrange(year) %>% 
  reframe(tibble(year2 = (min(year) + 1):(max(year) - 1),
                 unobserved = as.integer(!year2 %in% year),
                 age = year2 - yearborn[1],
                 age10 = (age - age_senesce) / 10)) %>% 
  ungroup() %>% 
  rename(year = year2) %>% 
  mutate(age_cat = factor(age >= age_senesce, 
                          labels = c("Young", "Old")),
         year_fct = factor(year)) %>% 
  left_join(select(sealdat, animalID, year, observed_int),
            by = c("animalID", "year"))

missing_mod <- glmer(unobserved ~ age10 + (1 | year_fct) + (1 | animalID), 
                     missing_seals, 
                     family = binomial)

@@

missing_age_se <- summary(missing_mod)$coef["age10", "Std. Error"]
missing_age_est <- summary(missing_mod)$coef["age10", "Estimate"]
missing_age_coef <- sprintf("%0.2f (95%% CI: %0.2f - %0.2f)",
                            missing_age_est,
                            missing_age_est - 1.96 * missing_age_se,
                            missing_age_est + 1.96 * missing_age_se)

missing_mod_old <- glmer(
  unobserved ~ age10 + (1 | year_fct),
  filter(missing_seals, age >= 11),
  family = binomial
)

summary(missing_mod_old)

missing_old_se <- summary(missing_mod_old)$coef["age10", "Std. Error"]
missing_old_est <- summary(missing_mod_old)$coef["age10", "Estimate"]
missing_old_coef <- sprintf("%0.2f (95%% CI: %0.2f - %0.2f)",
                            missing_old_est,
                            missing_old_est - 1.96 * missing_old_se,
                            missing_old_est + 1.96 * missing_old_se)

@@

First, we tested whether probability of detection was related to age. We found the proportion of unobserved animals increased with age. For this analysis, we excluded the first and last year of observation for each animal, because animals were necessarily observed in those years. For all years between, we determined whether animals were observed. Then we estimated the effect of age on probability of an animal being unobserved using a GLMM (binomial response, logit link) with random effects for year and animal ([@fig-unobserved]). The coefficient for age was `r missing_age_coef` (*p*\<0.0001). The old-animals-only model estimated the age coefficient to be `r missing_old_coef`.

```{r}
#| label: fig-unobserved
#| fig-cap: The probability of an animal being unobserved increased with age. Points and error bars represent the proportion of unobserved animals within age classes plus/minus one standard error. The blue line and shaded area are the mean and 95% CI probability of an animal being unobserved estimated with a generalized linear mixed effect model. See text for model details.

missing_grid <- expand_grid(age = 5:19, 
                            year_fct = factor(1989:2022)) %>% 
  mutate(age10 = (age - age_senesce) / 10)

missing_pred <- missing_grid %>% 
  mutate(predicted = predict(missing_mod, 
                             newdata = .,
                             re.form = ~ (1 | year_fct)),
         se = suppressWarnings(predict(missing_mod, 
                                       newdata = .,
                                       se.fit = TRUE,
                                       re.form = NA)$se.fit),
         conf.low = predicted - 1.96 * se,
         conf.high = predicted + 1.96 * se) %>% 
  left_join(count(missing_seals, year_fct), by = "year_fct") %>% 
  group_by(age) %>% 
  summarize(across(c(predicted, conf.low, conf.high), mean)) %>% 
  mutate(across(c(predicted, conf.low, conf.high), family(missing_mod)$linkinv))

missing_pred_years <- missing_grid %>% 
  mutate(predicted = predict(missing_mod, 
                             newdata = .,
                             type = "response",
                             re.form = ~ (1 | year_fct)))

missing_pred_old <- expand_grid(age = 11:19, 
                                year_fct = factor(1989:2022)) %>% 
  mutate(age10 = (age - age_senesce) / 10) %>% 
  mutate(predicted = predict(missing_mod_old, 
                             newdata = .,
                             re.form = ~ (1 | year_fct)),
         se = suppressWarnings(predict(missing_mod, 
                                       newdata = .,
                                       se.fit = TRUE,
                                       re.form = NA)$se.fit),
         conf.low = predicted - 1.96 * se,
         conf.high = predicted + 1.96 * se) %>% 
  left_join(count(filter(missing_seals, age >= 11), year_fct), 
            by = "year_fct") %>% 
  group_by(age) %>% 
  summarize(across(c(predicted, conf.low, conf.high), mean)) %>% 
  mutate(across(c(predicted, conf.low, conf.high), family(missing_mod)$linkinv))

missing_pred_pop <- tibble(age = 5:19,
                           age10 = (age - age_senesce) / 10) %>% 
  mutate(predicted = predict(missing_mod, 
                             newdata = .,
                             type = "response",
                             re.form = NA))

missing_seals %>% 
  group_by(age) %>% 
  summarize(n = n(),
            n_unobserved = sum(unobserved),
            unobserved_mean = n_unobserved / n,
            unobserved_lwr = binom.test(n_unobserved, n)$conf.int[1],
            unobserved_upr = binom.test(n_unobserved, n)$conf.int[2]) %>% 
  ggplot(aes(age)) +
  geom_line(aes(y = predicted, group = year_fct), 
            missing_pred_years, 
            alpha = 0.2) +
  # Overall model
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              missing_pred,
              alpha = 0.3,
              fill = "firebrick") +
  geom_line(aes(y = predicted),
            missing_pred,
            color = "firebrick",
            linewidth = 2) +
  # Old only
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              missing_pred_old,
              alpha = 0.3,
              fill = "cornflowerblue") +
  geom_line(aes(y = predicted),
            missing_pred_old,
            linetype = "dashed",
            color = "cornflowerblue",
            linewidth = 2) +
  # Raw data
  geom_pointrange(aes(y = unobserved_mean, 
                      ymin = unobserved_lwr, ymax = unobserved_upr)) +
  scale_x_continuous("Age", breaks = seq(5, 19, by = 4)) +
  scale_y_continuous("Unobserved proportion", 
                     labels = scales::label_percent(accuracy = 1),
                     limits = c(0, NA))
```

@@

@@

TODO: AIC comparison

@@

# H2b

recr_age_quad <- glmer(
  pup_recr_int ~ poly(age10, 2, raw = TRUE) + (1 | year_fct),
  filter(surv_recr_data, pupsex == "F"),
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

@@

>>>>

@@

@@

## Are we missing animals more or less with age?
missing_seals <- sealdat %>% 
  filter(age <= 20) %>% 
  # Fill in non-resighted years
  group_by(animalID) %>% 
  arrange(year) %>% 
  reframe(tibble(year2 = (yearborn[1] + 4):max(year), # from age 4 to last obs
                 unobserved = as.integer(!year2 %in% year),
                 age = year2 - yearborn[1],
                 age10 = (age - age_senesce) / 10)) %>% 
  ungroup() %>% 
  rename(year = year2) %>% 
  filter(year >= 1988) %>% 
  mutate(age_cat = factor(age >= age_senesce, 
                          labels = c("Young", "Old")),
         year_fct = factor(year)) %>% 
  left_join(select(sealdat, animalID, year, observed_int),
            by = c("animalID", "year"))

# Summary: mean and 95% CI 
missing_summ <- missing_seals %>% 
  group_by(age, age_cat) %>% 
  summarize(n = n(),
            n_unobserved = sum(unobserved),
            unobserved_mean = n_unobserved / n,
            unobserved_lwr = binom.test(n_unobserved, n)$conf.int[1],
            unobserved_upr = binom.test(n_unobserved, n)$conf.int[2],
            .groups = "drop")

missing_mod <- glmer(
  unobserved ~ age10 + (1 | year_fct) + (1 | animalID), 
  missing_seals, 
  family = binomial
)

@@

@@

First, we tested whether probability of detection was related to age. We found the proportion of unobserved animals decreased with age (i.e., older seals are more likely to be observed). For this analysis, we excluded the last year each animal was observed. Then, for all years between animals' age 4 year to the year prior to final observation, we determined whether each animal was observed. We estimated the effect of age on probability of an animal being unobserved using a GLMM (binomial response, logit link) with random effects for year and animal ([@fig-unobserved]). The coefficient for age was -0.2593 (95% CI: -0.4896 - -0.0290, *p* = 0.027). The change in mean unobserved rate from age 4 to 20 was 20.0% to 14.2%.

```{r}
#| label: fig-unobserved
#| fig-cap: The probability of an animal being unobserved decreased with age. Points and error bars represent the proportion of unobserved animals with 95% CI. The blue line and shaded area are the mean and 95% CI probability of an animal being unobserved (GLMM, with random effect of year weighted by number of seals observed each year). The dotted line is the mean response of the unweighted predictions. Numbers at the top indicate number of seals in each age category. See text for model details.

# Prediction grid
missing_grid <- expand_grid(age = 4:20, 
                            year_fct = factor(1989:2022)) %>% 
  mutate(age10 = (age - age_senesce) / 10)

# Missing seal predictions (weighted model)
missing_pred <- missing_grid %>% 
  mutate(predicted = predict(missing_mod, 
                             newdata = .,
                             re.form = ~ (1 | year_fct)),
         pred_pop = predict(missing_mod, 
                            newdata = .,
                            type = "response",
                            re.form = NA),
         se = suppressWarnings(predict(missing_mod, 
                                       newdata = .,
                                       se.fit = TRUE,
                                       re.form = NA)$se.fit),
         conf.low = predicted - 1.96 * se,
         conf.high = predicted + 1.96 * se) %>% 
  left_join(count(missing_seals, year_fct), by = "year_fct") %>% 
  group_by(age) %>% 
  summarize(across(c(predicted, pred_pop, conf.low, conf.high), mean)) %>% 
  mutate(across(c(predicted, conf.low, conf.high), 
                family(missing_mod)$linkinv))

# Yearly predictions
missing_pred_years <- missing_grid %>% 
  mutate(predicted = predict(missing_mod, 
                             newdata = .,
                             type = "response",
                             re.form = ~ (1 | year_fct)))

ggplot(missing_pred, aes(age)) +
  # Line per year
  geom_line(aes(y = predicted, group = year_fct), 
            missing_pred_years, 
            alpha = 0.2) +
  # Weighted model
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.3,
              fill = "cornflowerblue") +
  geom_line(aes(y = predicted),
            linewidth = 2,
            color = "cornflowerblue") +
  # Population-level model
  geom_line(aes(y = pred_pop), 
            linetype = "dotted",
            linewidth = 1.2,
            color = "cornflowerblue") +
  # Raw data
  geom_pointrange(aes(y = unobserved_mean, 
                      ymin = unobserved_lwr, 
                      ymax = unobserved_upr),
                  missing_summ) +
  # Sample size
  geom_text(aes(age, label = n), missing_summ, y = 0.8, vjust = 1.1) +
  scale_x_continuous("Age", breaks = seq(4, 20, by = 4)) +
  scale_y_continuous("Unobserved proportion", 
                     labels = scales::label_percent(accuracy = 1),
                     limits = c(0, NA)) +
  theme(legend.position = "none")
```

@@

This needs to be rethought now!

@@

@@

# H2b
recr_age_quad <- glmer(
  pup_recr_int ~ poly(age10, 2, raw = TRUE) + (1 | year_fct),
  filter(surv_recr_data, pupsex == "F"),
  family = "binomial",
  control = glmerControl(optimizer = "bobyqa")
)

@@

### AIC Comparison

```{r}
#| label: quad-aic

aic <- function(break_mod, quad_mod, hyp) {
  AIC(break_mod, quad_mod) %>% 
    as_tibble() %>% 
    mutate(dAIC = AIC - min(AIC),
         rellik = exp(-0.5 * dAIC),
         weight = rellik / sum(rellik),
         hypothesis = hyp,
         model = c("Breakpoint", "Quadratic")) %>% 
  select(hypothesis, model, AIC, weight)
}

pmap(list(break_mod = list(breed_age_mod, pup_surv_mod, pup_recr_mod),
          quad_mod = list(breed_age_quad, surv_age_quad, recr_age_quad),
          hyp = c("H1", "H2a", "H2b")),
     aic) %>% 
  list_rbind() %>% 
  knitr::kable()
```
